Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-28 paper.tex
ENH:  Submission ready.

\begin{abstract}
Diffeomorphic maps are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility 
taking the form of physics-based constraints or through some approximation thereof, e.g. 
Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In 
the context of the original Demons' framework, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}
can be viewed as a smoothing alternative  in which explicit regularization is achieved through 
fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image 
registration solutions with several advantages.  Implementation 
is open source and available through the Insight Toolkit and our Advanced Normalization Tools
(ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm 
\citep{avants2008}, implemented within the same framework, and its B-spline analog is 
performed using open labeled brain data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Variants of three popular diffeomorphic algorithms and their DMFFD
analogs (LDDMM, DARTEL, and SyN) were implemented by the authors as
part of the recent refactoring of the open source Insight Toolkit
(ITK) although related work had been previously implemented within the
popular Advanced Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the greedy variant
of SyN, our evaluation focus in this work is its B-spline
analog which we denote as ``B-spline SyN'' or ``DMFFD SyN.''
Evaluations of the respective algorithmic
instantiations are performed using the \verb#antsRegistration# program
found in the ANTs repository (also originally developed by the
authors).  This permits a direct algorithmic comparison as potential
sources for implementation bias have been reduced
\citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and scripts to reproduce the results contained in
this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible). 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Note that any pair of images can be specified per metric per stage.

@@

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
The perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
It should be noted that for the
data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.

@@

B-spline SyN produced slightly greater Dice values than the original SyN. 
Although actual differences are relatively small, they are statistically 
significant. By implementing both algorithms in the the same code base, 
we are not only able to eliminate non-regularization components of the 
registration but we are also able to eliminate implementation differences. 
Thus, performance disparity can be isolated to smoothing choice. However, 
even with this restricted focus there are various reasons for the evaluation 
outcome. These include the approximation-versus-convolution distinction 
mentioned earlier for the two regularization approaches (which could also 
explain the reason why the range of log Jacobian values tend to be significantly 
higher for B-spline SyN). Also, the fact that the regularization for 
B-spline SyN is theoretically continuous whereas the truncated Gaussian 
convolution is only a discrete approximation could be a potential factor. 
Exploration of other possibilities are planned for future work.

One of the advantages that has not been explored in this manuscript is 
the use of B-spline SyN for small deformation estimation problems such 
as in pulmonary or cardiac applications.  Such problems typically require
greater regularization which implies larger discrete kernels for 
Gaussian convolution.
A related issue is applications involving severely anisotropic data where
the continuous nature of the DMFFD approach might help over Gaussian convolution.
Additionally, we have not yet explored the use of sparsely sampled data 
and the advantages of the proposed framework.
Ongoing work will continue to explore these issues.

@@

A significant amount of research has been devoted to image registration
algorithmic development.  Given their many salient characteristics particularly with
respect to large deformation estimation constrained by topological 
continuity, diffeomorphic registration
approaches have been a particular focus in the neuroimaging community.
However, many groups continue to find success with non-diffeomorphic
FFD methods
\citep[e.g.][]{rueckert1999,klein2010}.  Using our DMFFD framework, 
B-spline regularization is easily adapted into the diffeomorphic registration 
framework and performs comparably to analogous algorithms which we demonstrated 
in this work for the case of the widely-used SyN.

>>>>

\begin{abstract}
Diffeomorphic mappings are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular diffeomorphic image registration algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility 
taking the form of physics-based constraints or through some approximation thereof, e.g. 
Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In 
the context of the original Demons' framework, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}
can be viewed as a smoothing alternative  in which explicit regularization is achieved through 
fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image 
registration solutions with several advantages.  Implementation 
is open source and available through the Insight Toolkit and our Advanced Normalization Tools
(ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm 
\citep{avants2008}, implemented within the same framework, and its B-spline analog is 
performed using open labeled brain data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Three popular diffeomorphic algorithms and their DMFFD
analogs (LDDMM, DARTEL, and SyN) were implemented by the authors as
part of the recent refactoring of the open source Insight Toolkit
(ITK) although related work had been previously implemented within the
popular Advanced Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the greedy variant
of SyN, our evaluation focus in this work is its B-spline
analog which we denote as ``B-spline SyN'' or ``DMFFD SyN.''
Evaluations of the respective algorithmic
instantiations are performed using the \verb#antsRegistration# program
found in the ANTs repository (also originally developed by the
authors).  This permits a direct algorithmic comparison as potential
sources for implementation bias have been reduced
\citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and scripts to reproduce the results contained in
this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible). 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Any pair of images can be specified per metric per stage.

@@

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
The perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
For the
data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.

@@

B-spline SyN produced slightly greater Dice values than the original SyN. 
Although actual differences are relatively small, they are statistically 
significant. By implementing both algorithms in the the same code base, 
we are not only able to eliminate non-regularization components of the 
registration but we are also able to eliminate implementation differences. 
Thus, performance disparity can be isolated to smoothing choice. However, 
even with this restricted focus there are various reasons for the evaluation 
outcome. These include the approximation-versus-convolution distinction 
mentioned earlier for the two regularization approaches (which could also 
explain the reason why the range of log Jacobian values tend to be significantly 
higher for B-spline SyN). Also, the fact that the regularization for 
B-spline SyN is theoretically continuous whereas the truncated Gaussian 
convolution is only a discrete approximation could be a potential factor. 
%Exploration of other possibilities are planned for future work.

Relative to Klein's study \citep{klein2009}, it should be noted that 
the SyN implementation used in that study is found in the {\tt ANTS} 
program which is a precursor of the {\tt antsRegistration} program
described earlier.  Although the theoretical aspects 
are the same, there are substantial implementation differences between 
the two programs.  In addition, several parameters varied between the two
studies which translated into a more aggressive metric and 
gradient step in addition to fewer levels and iterations.%
\footnote{
Specifically, as provided in the supplementary material of \cite{klein2009},
the command line parameters were:
{\tt ANTS 3 -m PR[\$\{fixed\},\$\{moving\},1,2] -o \$\{output\} -r Gauss[2,0] 
-t SyN[0.5] -i 30x99x11 --use-Histogram-Matching}.  
}
In this study we took a more conservative approach based on our 
continued development and experience resulting in parameters which have proven
useful in our cortical thickness pipeline (encapsulated in 
the ANTs script {\tt antsCorticalThickness.sh}) and our
experience with the MICCAI 2013 SATA challenge data.  We also 
ran our own internal experiments with Gaussian SyN (both {\tt ANTS} and 
{\tt antsRegistration}) using the same conservative parameters 
on the MAL35 data for which the latter demonstrated slightly improved
performance over the former.

One of the advantages that has not been explored in this work is 
the use of B-spline SyN for small deformation estimation problems such 
as in pulmonary or cardiac applications.  Such problems typically require
greater regularization which implies larger discrete kernels for 
Gaussian convolution.
A related issue is applications involving severely anisotropic data where
the continuous nature of the DMFFD approach might help over Gaussian convolution.
Additionally, we have not yet explored the use of sparsely sampled data 
and the advantages of the proposed framework.
Ongoing work will continue to explore these issues.

@@
